### 2024-07-10 causal-inference.qmd
20240710 - tweaks

In general, to meet the third criterion of causality, one must consider the counterfactual.
A *counterfactual* is a hypothetical scenario—it did not occur.
However, when engaging in causal inference, it is important to consider what would have happened had the hypothetical scenario occurred.
For instance, consider that we conduct an experiment to randomly assign some players to consume a sports drink before a game and other players to drink only water.
In this case, our treatment/intervention group is the group of players that consumed a sports drink.
The control group is the group players that drank only water.
Now, consider that the players in the treatment group outperform the players in the the control group in their football game.
In such a study, we observe what *did happen* when players received a treatment.
The counterfactual is knowledge of what *would have happened* to those same players if they simultaneously had not received treatment [@Shadish2002].
The true causal effect, then, is the difference between what did happen and what would have happened.
However, we cannot observe a counterfactual.
That is, we do not know for sure what would have happened to the players who received treatment if those same players had actually not received treatment.
We have a control group, but the control group does not have the same players as the intervention group, and it is impossible for a person to simultaneously receive and not receive treatment.

So, our goal in working toward causal inference as scientists is to create reasonable approximations to this impossible counterfactual [@Shadish2002].
For instance, if using a [between-subject design](#sec-withinSubject), we want the two groups to be equivalent in every possible way except whether or not they receive the treatment, so we might stratify each group to be equivalent in terms of age, weight, position, experience, skill, etc.
Or, we might test the same people using an A-B-A-B [within-subject design](#sec-withinSubject).
In an A-B-A-B [within-subject design](#sec-withinSubject), players receive no treatment at baseline (timepoint 1: game 1), receive the treatment at timepoint 2 (game 2), receive no treatment at timepoint 3 (game 3), and receive the treatment at timepoint 4 (game 4).
Neither of these approximations is a true counterfactual.
In the [between-subject design](#sec-withinSubject), the players differ between the two groups, so we cannot know how the individuals who received the treatment would have performed if they had actually not received the treatment.
In the A-B-A-B [within-subject design](#sec-withinSubject), the players are the same, but they timepoints that they receive or do not receive the treatment differ, and there can be [carryover effects](#sec-withinSubject) from one condition to the next.
For instance, consuming sports drinks before game 2 might also help them be better hydrated in general, including, for subsequent games.
Thus, we cannot know how a player would have performed in game 1 with treatment or in game 2 without treatment, etc.
Nevertheless, it is important to be aware of the counterfactual and to engage in counterfactual reasoning to consider what would have happened if various hypothetical scenarios had occurred.
Considering the counterfactual is important for designing closer approximations to the counterfactual in studies for stronger research designs and stronger causal inference.

>>>>

In general, to meet the third criterion of causality, one must consider the counterfactual.
A *counterfactual* is what would have happened in the hypothetical scenario that the cause did not occur [i.e., what would have happened in the absence of the cause; @Shadish2002].
When engaging in causal inference, it is important to consider what would have happened if the hypothetical cause had actually not occurred.
For instance, consider that we conduct an experiment to randomly assign some players to consume a sports drink before a game and other players to drink only water.
In this case, our treatment/intervention group is the group of players that consumed a sports drink.
The control group is the group players that drank only water.
Now, consider that the players in the treatment group outperform the players in the the control group in their football game.
In such a study, we observe what *did happen* when players received a treatment.
The counterfactual is knowledge of what *would have happened* to those same players if they simultaneously had not received treatment [@Shadish2002].
The true causal effect, then, is the difference between what did happen and what would have happened.
However, we cannot observe a counterfactual.
That is, we do not know for sure what would have happened to the players who received treatment if those same players had actually not received treatment.
We have a control group, but the control group does not have the same players as the intervention group, and it is impossible for a person to simultaneously receive and not receive treatment.

So, our goal in working toward causal inference as scientists is to create reasonable approximations to this impossible counterfactual [@Shadish2002].
For instance, if using a [between-subject design](#sec-withinSubject), we want the two groups to be equivalent in every possible way except whether or not they receive the treatment, so we might stratify each group to be equivalent in terms of age, weight, position, experience, skill, etc.
Or, we might test the same people using an A-B-A-B [within-subject design](#sec-withinSubject).
In an A-B-A-B [within-subject design](#sec-withinSubject), players receive no treatment at baseline (timepoint 1: game 1), receive the treatment at timepoint 2 (game 2), receive no treatment at timepoint 3 (game 3), and receive the treatment at timepoint 4 (game 4).
Neither of these approximations is a true counterfactual.
In the [between-subject design](#sec-withinSubject), the players differ between the two groups, so we cannot know how the individuals who received the treatment would have performed if they had actually not received the treatment.
In the A-B-A-B [within-subject design](#sec-withinSubject), the players are the same, but they timepoints that they receive or do not receive the treatment differ, and there can be [carryover effects](#sec-withinSubject) from one condition to the next.
For instance, consuming sports drinks before game 2 might also help them be better hydrated in general, including, for subsequent games.
Thus, we cannot know how a player would have performed in game 1 with treatment or in game 2 without treatment, etc.
Nevertheless, it is important to be aware of the counterfactual and to engage in counterfactual reasoning to consider what would have happened if the supposed cause had occurred.
Considering the counterfactual is important for designing closer approximations to the counterfactual in studies for stronger research designs and stronger causal inference.
